The MACE is essentially unidimensional: Bifactor models of the `Maltreatment and Abuse Chronology of Exposure' scale

### 2022-08-07 manuscript/main.tex
manuscipt edits (2022-08-07a)

All participants completed the 52-item version of the MACE \citep{teicher2015maltreatment}. The scale is a retrospective self-report measure of the severity and timing of exposure to 10 types of childhood maltreatment. For each item, participants endorse whether they ever experienced a particular event during childhood (e.g. ``Parents or guardians intentionally pushed, grabbed, shoved, slapped, pinched, punched or kicked you'') and, if so, at what ages (up to age 18). Here we only analyze the dichotomous endorsement responses (Yes = 1, No = 0) and not the chronology data. Six of the 52-items --- evenly divided between the emotional and physical neglect subscales --- are positively valenced and reverse-scored (e.g. ``One or more individuals in your family helped you feel important or special''). For these items we follow the scoring convention set by \cite{teicher2015maltreatment}: a score of 1 is assigned to a response only in the absence of a positive event for all 18 years of childhood. This is in contrast to the remaining items in the scale, where a score of 1 is assigned to a response if maltreatment or abuse occurred in at least one year of childhood.

% (i.e. verbal abuse, physical abuse, nonverbal emotional abuse, sexual abuse, emotional neglect, and physical neglect by parents or guardians)

@@

Using the cutoffs recommended by \cite{hidalgo2014binary}, we identified 11 of 52 items on the MACE that exhibited large DIF by study (Figure \ref{fig:dif_study}). Interestingly these included all six reverse-scored items. We also found an additional 11 items that exhibited large DIF by gender (Figure \ref{fig:dif_gender}). Five of these items measured sexual abuse (endorsed by more women than men), and another five measured peer physical abuse (endorsed by more men than women). We identified no items that exhibited large DIF by age. Given the number of items exhibiting DIF by sample membership, we decided to analyze each sample separately.

@@

To investigate the factor structure of the MACE, we fit a series of confirmatory item factor models separately to the response data from each sample. The basis of each model is the graded response model for ordered polytomous data \citep{samejima1997graded}. The factor structure for each of the four models is depicted in Figure \ref{fig:models}. The first model is a one-factor (undimensional) model where all items load onto a general maltreatment factor (Figure \ref{fig:models}a). The second is a bifactor model where all items load onto a general maltreatment factor and also one of 10 specific factors based on their subscale membership (Figure \ref{fig:models}b).

The third model we fit investigated a conceptual model of maltreatment that distinguishes between threat and deprivation as distinct forms of childhood adversity \citep{mclaughlin2014childhood}. Initially we fit a bifactor model where all items loaded onto a general maltreatment factor and either the threat or deprivation specific factors. However, we observed vanishing item loadings on the threat specific factor. As such, we instead fit a bifactor S-1 model \citep{eid2017anomalous} where all items load onto a general maltreatment (or reference) factor, and only the items measuring neglect experiences load onto their own specific factor (Figure \ref{fig:models}c).

@@

The fourth and final model proposes a more parsimonious factor structure for the MACE. Items in the MACE can be neatly divided according to the source of adversity: maltreatment by adults at home (e.g. parents) or by peers. We hypothesized then that there would be two principal factors underlying responses on the MACE, maltreatment by adults and peer victimization. Based on preliminary model fits, we also suspected the presence of a secondary methods factor affecting the 6 reverse-scored items. Initially we fit a bifactor model in which all items loaded onto a general maltreatment factor and also one of three specific factors: maltreatment by adults, peer victimization, or reverse-scoring. However, we also observed vanishing loadings for the maltreatment by adults specific factor. Thus, we instead fit a bifactor S-1 model where all items load onto a general maltreatment factor; in addition, the peer victimization and reverse-scoring items load onto their own respective specific factors (Figure \ref{fig:models}d).

@@

Given the limitations of SEM-based fit indices for item response models (e.g. \citealt{reise2014evaluating}), we calculated two additional fit indices. First, we used posterior predictive model checking to calculate a $\chi^2_{NC}$ discrepancy measure based on the total score distribution \citep{sinharay2006posterior}. This measure compares the observed and model-predicted proportion of participants at each total score level. Second, we calculated Yen's $Q_3$ statistic for each pair of items to detect local dependence \citep{yen1984effects} and also the critical $Q_3$-value above which local dependence is indicated  \citep{christensen2017critical}. Briefly, for each model and sample, we simulated 1000 locally independent datasets using the posterior distribution of the model parameters. We then recorded the maximum $Q_3$ value per simulation. The critical $Q_3$-value was defined as the 99th percentile of these max $Q_3$ values.

Finally, the goodness-of-fit of the models was compared using Bayesian leave-one out cross-validation (LOO-CV; \citealt{vehtari2017practical}). The LOO-CV measure quantifies the discrepancy between the model and the data while taking into account model complexity. Here LOO-CV values are presented in deviance scale, such that smaller values indicate better fit.

@@

To evaluate the reliability of the MACE total and subscale scores, we computed several model-based reliability indices using the factor loading estimates from the bifactor models. First, we calculated coefficient omega ($\omega$; \citealt{mcdonald1999test}), which quantifies the proportion of reliable score variance attributable to all factors (i.e. both group and specific factors). Next we calculated coefficient omega hierarchical ($\omega_h$) and coefficient omega subscale ($\omega_s$;  \citealt{reise2013scoring}). Whereas $\omega_h$ measures the fraction of total score variance attributable to the general factor, $\omega_s$ measures the fraction of subscale score variance attributable to a specific factor after controlling for the general factor. The values of $\omega_h$/$\omega_s$ facilitate interpretation of total and subscale scores. Values of $\omega_h$ approaching 1 indicate the general factor is the primary source of reliable variance in a total score. When $\omega_h$ > 0.80, the total score may be considered an essentially unidimensional reflection of the general factor even if the data are multidimensional \citep{rodriguez2016applying}. So too values of $\omega_s$ approaching 1 indicate the specific factor, and not the general factor, is the primary source of reliable variance in a subscale score. \cite{canivez2016bifactor} suggested that an acceptable $\omega_s$ value is 0.50 and that >0.70 is desirable. In contrast, values of $\omega_s$ smaller than 0.50 preclude the interpretation of subscale scores as reflecting a specific factor \citep{gignac2013bifactor}. 
To assess the unidimensionality of the MACE, we also calculated the explained common variance (ECV; \citealt{sijtsma2009use}) and the proportion of uncontaminated correlations (PUC; \citealt{reise2013multidimensionality}). The ECV is the ratio of the variance explained by the general factor divided by the total common variance of a scale; it therefore measures the contribution of the general factor relative to the specific factors. As ECV values approach 1, the general factor loadings of a bifactor model are expected to resemble those that would be obtained by a one-dimensional model. ECV values greater than 0.70 often indicate a scale is essentially unidimensional \citep{rodriguez2016applying}. In turn, the PUC quantifies how many inter-item correlations are accounted for only by the general factor and is an important moderator of the ECV. As the PUC increases, ECV is less important for evaluating the risk of bias when fitting a unidimensional model to data with a latent bifactor structure. When PUC > 0.80, there is a low risk of bias when a multidimensional scale is treated as unidimensional \citep{reise2013multidimensionality}. As an additional check, we calculated the relative parameter bias (RPB), which is the difference between an item's loading in a unidimensional model and its corresponding general factor loading the bifactor, divided by the general factor loading in the bifactor. Parameter bias less than 10–15\% is acceptable \citep{muthen1987structural}.

>>>>

All participants completed the 52-item version of the MACE \citep{teicher2015maltreatment}. The MACE is a retrospective self-report scale that measures 10 unique types of childhood maltreatment (i.e. parental verbal abuse, physical abuse, nonverbal emotional abuse, sexual abuse, emotional neglect, physical neglect, witnessing violence to siblings, witnessing interparental violence, peer verbal abuse, peer physical abuse). For each item, participants endorse whether they ever experienced a particular event during childhood (e.g. ``Parents or guardians intentionally pushed, grabbed, shoved, slapped, pinched, punched or kicked you'') and, if so, at what ages (up to age 18). Here we only analyze the dichotomous endorsement responses (Yes = 1, No = 0) and not the chronology data. Six of the 52-items --- evenly divided between the emotional and physical neglect subscales --- are positively valenced and reverse-scored (e.g. ``One or more individuals in your family helped you feel important or special''). For these items we follow the scoring convention set by \cite{teicher2015maltreatment}: a score of 1 is assigned to a response only in the absence of a positive event for all 18 years of childhood. This is in contrast to the remaining items in the scale, where a score of 1 is assigned to a response if maltreatment or abuse occurred in at least one year of childhood.

@@

Using the cutoffs recommended by \cite{hidalgo2014binary}, we identified 11 of 52 items on the MACE with large DIF by study (Figure \ref{fig:dif_study}). These included all six reverse-scored items. We also found an additional 11 items with large DIF by gender (Figure \ref{fig:dif_gender}). Five of these items measured sexual abuse (endorsed more frequently by women), and another five measured peer physical abuse (endorsed more frequently by men). We identified no items that exhibited large DIF by age. Given the number of items with DIF by sample membership, we decided to analyze each sample separately.

@@

To investigate the factor structure of the MACE, we fit a series of confirmatory item factor models separately to the response data from each sample. The basis of each model is the graded response model for ordered polytomous data \citep{samejima1997graded}. The factor structure for each of the four models is depicted in Figure \ref{fig:models}. The first model is a one-factor (undimensional) model where all items load onto a general maltreatment factor (Figure \ref{fig:models}a). The second is a bifactor model where all items load onto a general maltreatment factor and also one of 10 specific factors based on their subscale membership (Figure \ref{fig:models}b). This model will be used to evaluate if the MACE can produce reliable subscale scores for each of the 10 types of maltreatment assessed by the MACE.

The third model was motivated by a theoretical account of childhood maltreatment that distinguishes between threat and deprivation experiences as distinct types of adversity \citep{mclaughlin2014childhood}. To explore this model of the MACE, we initially fit a bifactor model where all items loaded onto a general maltreatment factor, and also a threat or neglect specific factor. However, we observed vanishing item loadings on the threat specific factor. As such, we instead fit a bifactor S-1 model \citep{eid2017anomalous} where all items load onto a general maltreatment factor, and only items measuring neglect load onto a specific factor (Figure \ref{fig:models}c). This model will be used to evaluate if the MACE can produce reliable neglect subscale scores distinguishable from general maltreatment.

@@

The fourth and final model proposes a more parsimonious structure of the MACE based on the content of its items. Items on the MACE can be organized according to the source of maltreatment: parents (or other adults in the home) and peers. Thus, we hypothesized there would be two factors underlying responses on the MACE: a primary maladaptive parenting factor and a secondary peer victimization factor. Based on preliminary model fitting, we also suspected the presence of a tertiary methods factor affecting the six reverse-scored items. We initially fit a bifactor model in which all items loaded onto a general maltreatment factor and one of three specific factors (maladaptive parenting, peer victimization, reverse-scoring); however, we observed vanishing loadings on the first specific factor. As such, we instead fit a bifactor S-1 model where all items load onto a general maltreatment factor, and the peer victimization \& reverse-scoring items additionally load onto their own specific factors (Figure \ref{fig:models}d). This model will be used to evaluate if the MACE can produce peer victimization reliable subscale scores distinguishable from general maltreatment.

@@

Given the limitations of SEM-based fit indices for item response models (e.g. \citealt{reise2014evaluating}), we calculated two additional fit indices. First, we used posterior predictive model checking to calculate a $\chi^2_{NC}$ discrepancy measure based on the total score distribution \citep{sinharay2006posterior}. This measure compares the observed and model-predicted proportion of participants at each total score level. Second, to test for local dependence in each pair of items, we calculated Yen's $Q_3$ statistic \citep{yen1984effects} and also the critical $Q_3$-value (i.e. the value above which local dependence is indicated; \citealt{christensen2017critical}). For each model and sample, we simulated 1000 locally independent datasets using the posterior distribution of the model parameters. We then recorded the maximum $Q_3$ value per simulation. The critical $Q_3$-value was defined as the 99th percentile of these max $Q_3$ values.

The goodness-of-fit of the models was compared using Bayesian leave-one out cross- validation (LOO-CV; \citealt{vehtari2017practical}). The LOO-CV measure quantifies the discrepancy between the model and data while taking into account model complexity. LOO-CV values are presented in deviance scale, i.e. smaller values indicate better fit.

@@

To evaluate the reliability of the MACE total and subscale scores, we computed several model-based reliability indices using the factor loading estimates from the bifactor models. First, we calculated coefficient omega ($\omega$; \citealt{mcdonald1999test}), which quantifies the proportion of reliable score variance attributable to all factors (i.e. both group and specific factors). Next we calculated coefficient omega hierarchical ($\omega_h$) and coefficient omega subscale ($\omega_s$;  \citealt{reise2013scoring}). Whereas $\omega_h$ measures the fraction of total score variance attributable to the general factor, $\omega_s$ measures the fraction of subscale score variance attributable to a specific factor after controlling for the general factor. The values of $\omega_h$/$\omega_s$ facilitate interpretation of total and subscale scores. Values of $\omega_h$ approaching 1 indicate the general factor is the primary source of reliable variance in a total score. When $\omega_h$ > 0.80, the total score may be considered an essentially unidimensional reflection of the general factor even if the data are multidimensional \citep{rodriguez2016applying}. So too values of $\omega_s$ approaching 1 indicate the specific factor, and not the general factor, is the primary source of reliable variance in a subscale score. \cite{canivez2016bifactor} suggested that an acceptable $\omega_s$ value is 0.50 and that >0.70 is desirable. In contrast, values of $\omega_s$ smaller than 0.50 preclude the interpretation of subscale scores as reflecting a specific factor \citep{gignac2013bifactor}.

To assess the unidimensionality of the MACE, we also calculated the explained common variance (ECV; \citealt{sijtsma2009use}) and the proportion of uncontaminated correlations (PUC; \citealt{reise2013multidimensionality}). The ECV is the ratio of the variance explained by the general factor divided by the total common variance of a scale; it therefore measures the contribution of the general factor relative to the specific factors. As ECV values approach 1, the general factor loadings of a bifactor model are expected to resemble those that would be obtained by a one-dimensional model. ECV values greater than 0.70 often indicate a scale is essentially unidimensional \citep{rodriguez2016applying}. In turn, the PUC quantifies how many inter-item correlations are accounted for only by the general factor and is an important moderator of the ECV. As the PUC increases, ECV is less important for evaluating the risk of bias when fitting a unidimensional model to data with a latent bifactor structure. When PUC > 0.80, there is a low risk of bias when a multidimensional scale is treated as unidimensional \citep{reise2013multidimensionality}. As an additional check, we calculated the relative parameter bias (RPB), which is the difference between an item's loading in a unidimensional model and its corresponding general factor loading the bifactor, divided by the general factor loading in the bifactor. Parameter bias less than 10–15\% is acceptable \citep{muthen1987structural}.
